---
title: "EDA"
author: "Akalanka Bandara"
date: "2025-02-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## EDA

<span style="color: blue; font-size: 15px;">After preparing the dataset(cleaning empty values and deleting unwanted columns then the new dataset is saved as "jan_full_prepared.csv").</span>


### loading the relevant libraries 
```{r libraries, results='hide', warning=FALSE}
library(tidyverse)
library(readr)
library(ggplot2)
library(skimr)
library(janitor)
library(lubridate)
library(ggplot2)
library(plotly)
```

### Importing the saved dataset

<span style="color: blue; font-size: 15px;">forcing the several columns to be relevent data types,</span>
 
```{r loading_dataset}
jan_final <- read_csv("../../data/january_full_prepared.csv", col_types = cols(
  rentalID = col_character(),  # Force rentalID to be character
  BikeID = col_character(),    # Force BikeID to be character
  start_station_number = col_character(),
  start_station = col_character(),
  bike_model = col_character(),
  total_duration_mins = col_double(),
  start_date = col_datetime(),
  s_date = col_date(),
  s_time = col_time()
))
attach(jan_final)
str(jan_final)
head(jan_final)


jan_final_full=read_csv("jan_final_full.csv", col_types = cols(
  rentalID=col_character(),
  s_date=col_date(),
  s_time=col_time(),
  start_station_number=col_character(),
  start_station=col_character(),
  e_date=col_date(),
  e_time=col_time(),
  end_station_number=col_character(),
  end_station=col_character(),
  BikeID=col_character(),
  bike_model=col_character(),
  total_duration_mins=col_double()
))
```



### UNIVARIANT ANALYSIS

<span style="color: black; font-size: 17px;">**1)rentalID**.</span>
```{r u1}
print(length(unique(jan_final$rentalID)))

```

<span style="color: red; font-size: 15px;">NOTE: since the rentalID is unique we can use it as a counting variable.</span>


<span style="color: black; font-size: 17px;">**2)start stations**.</span>
```{r 2}
print(length(unique(jan_final$start_station_number)))
print(length(unique(jan_final$start_station)))
```

<span style="color: blue; font-size: 15px;">below map represents start stations.</span>
```{r 3}
library(leaflet)
map_locations=read_csv("Stations_w_lat_lon.csv")
head(map_locations)


leaflet(map_locations) %>%
  addTiles() %>%
  addCircleMarkers(
    ~longitude, ~latitude,
    color = "blue",
    radius = 0.5,  # Adjust point size
    fillOpacity = 0.7
  )
```



<span style="color: blue; font-size: 15px;">since station is categorical variable we draw a barchart and see how the distribution and trending.</span>


```{r 4}
start_date_data=jan_final %>% 
  group_by(start_station_number) %>% 
  select(start_station_number) %>% 
  summarise(values=n())

head(start_date_data)
mean(start_date_data$values)


g1=ggplot(start_date_data,aes(x=start_station_number,y=values,))+
  geom_bar(stat="identity",width = 0.7,just = 0.5,fill="blue")+
  labs(title="no of rented bikes of each station",x="start station ID",y="no of rented bikes")+
  geom_smooth(aes(group = 1), method = "loess", se = FALSE, color = "red", linetype = "dashed")
g1
```


<span style="color: blue; font-size: 15px;"> made a function to identify the top 5 station rental count over the january month.</span>

```{r 5,results='hide', warning=FALSE}
top_station_data_func <- function(data, station_number) {
  data %>%
    filter(start_station_number == station_number) %>%
    group_by(s_date) %>%
    summarise(vals = n(), .groups = 'drop') %>%
    arrange(desc(vals)) %>%
    mutate(
      weekday_or_weekend = ifelse(wday(s_date) %in% c(1, 7), "Weekend", "Weekday")
    )
}

station_numbers <- c("001072", "001011", "002587", "001075", "002696")  

# Process data for each station
station_data_list <- lapply(station_numbers, function(station) {
  top_station_data_func(jan_final, station)
})

# Combine the list of data frames into one data frame with an additional column for station number
station_data = bind_rows(station_data_list, .id = "station_number")
station_data$station_number = factor(station_data$station_number, labels = station_numbers)
head(station_data)


```

<span style="color: blue; font-size: 15px;"> from the function we can identify the total rental count for top 5 stations over the month.</span>

```{r 6}
ggplot(station_data, aes(x = s_date, y = vals, fill = weekday_or_weekend)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ station_number, scales = "free") +
  labs(title = "Count of Start Stations Over Date", x = "Date", y = "Rented Bikes") +
  theme_minimal()
```

<span style="color: blue; font-size: 15px;"> as we can see weekdays have the most count .</span>

<span style="color: blue; font-size: 15px;"> bike rentals by date and start station(top 10 stations and top 10 days)</span>

```{r 7}
top10_days=as.Date(c("2023-01-31" ,"2023-01-24" ,"2023-01-11", "2023-01-19" ,"2023-01-18" ,"2023-01-30", "2023-01-26", "2023-01-17","2023-01-12","2023-01-09"))
top10_stations=c("001072","001011","002587","001075","002696","300083","000960","001064","001142","022179")

top10_count_station_and_top10_days_and_count=jan_final %>% 
  filter(start_station_number 
         %in% c("001072","001011","002587","001075","002696","300083","000960","001064","001142","022179"),
         s_date %in% as.Date(c("2023-01-31" ,"2023-01-24" ,"2023-01-11", "2023-01-19" ,"2023-01-18" ,"2023-01-30", "2023-01-26", "2023-01-17","2023-01-12","2023-01-09"))
  ) %>% 
  select(s_date,start_station_number) %>% 
  group_by(start_station_number,s_date) %>% 
  summarise(count=n())
#head(top10_count_station_and_top10_days_and_count)

ggplot(top10_count_station_and_top10_days_and_count,
       aes(x = s_date, y = start_station_number, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "green") +
  labs(title = "Bike Rentals by Date and Start Station (Top 10 Stations and top10 days)",
       x = "Date",
       y = "Start Station",
       fill = "count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


<span style="color: blue; font-size: 15px;"> testing top 5 stations renatal count for each station's top 6 days over hour .</span>


<span style="color: blue; font-size: 15px;"> 1)station 001072</span>

```{r 8}
head(station_data %>% filter(station_number=="001072") %>% arrange(desc(vals)))
max_s_hour_data_for_max_date1=jan_final %>% 
  filter(start_station_number=="001072",s_date %in% as.Date(c("2023-01-25","2023-01-10","2023-01-31","2023-01-11","2023-01-18","2023-01-12"))) %>%
  select(s_date,s_date,s_time,start_station_number) %>% 
  group_by(s_date,hour=hour(s_time)) %>% summarise(vals=n()) %>% arrange(s_date,desc(vals))

#head(max_s_hour_data_for_max_date1)

ggplot(max_s_hour_data_for_max_date1, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ s_date, ncol = 2) +  # Facet by date, with 2 columns
  labs(
    title = "Count of Start Station 001072 Over Hours for Multiple Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()
```

```{r}

max_s_hour_data_for_max_date1_but1=jan_final_full%>% 
  filter(end_station_number=="001072",e_date %in% as.Date(c("2023-01-25","2023-01-10","2023-01-31","2023-01-11","2023-01-18","2023-01-12"))) %>%
  select(e_date,e_time,end_station_number) %>% 
  group_by(e_date,hour=hour(e_time)) %>% summarise(vals=n()) %>% arrange(e_date,desc(vals))

ggplot(max_s_hour_data_for_max_date1_but1, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ e_date, ncol = 2) + 
  labs(
    title = "Count of end Station 001072 Over Hours for above peak Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()
```



<span style="color: blue; font-size: 15px;"> 2)station 001011</span>

```{r 9}
head(station_data %>% filter(station_number=="001011") %>% arrange(desc(vals)))
max_s_hour_data_for_max_date2=jan_final_full%>% 
  filter(start_station_number=="001011",s_date %in% as.Date(c("2023-01-17","2023-01-19","2023-01-11","2023-01-12","2023-01-18","2023-01-25"))) %>%
  select(s_date,s_date,s_time,start_station_number) %>% 
  group_by(s_date,hour=hour(s_time)) %>% summarise(vals=n()) %>% arrange(s_date,desc(vals))

head(max_s_hour_data_for_max_date2)

sum(max_s_hour_data_for_max_date2$vals)

ggplot(max_s_hour_data_for_max_date2, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ s_date, ncol = 2) + 
  labs(
    title = "Count of Start Station 001011 Over Hours for Multiple Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()
```
```{r}

max_s_hour_data_for_max_date2_but2=jan_final_full%>% 
  filter(end_station_number=="001011",e_date %in% as.Date(c("2023-01-17","2023-01-19","2023-01-11","2023-01-12","2023-01-18","2023-01-25"))) %>%
  select(e_date,e_time,end_station_number) %>% 
  group_by(e_date,hour=hour(e_time)) %>% summarise(vals=n()) %>% arrange(e_date,desc(vals))

ggplot(max_s_hour_data_for_max_date2_but2, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ e_date, ncol = 2) + 
  labs(
    title = "Count of end Station 001011 Over Hours for above peak Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()

```



<span style="color: blue; font-size: 15px;"> 3)station 002587</span>

```{r 10}
head(station_data %>% filter(station_number=="002587") %>% arrange(desc(vals)))
max_s_hour_data_for_max_date3=jan_final_full%>% 
  filter(start_station_number=="002587",s_date %in% as.Date(c("2023-01-31","2023-01-12","2023-01-19","2023-01-26","2023-01-09","2023-01-11"))) %>%
  select(s_date,s_date,s_time,start_station_number) %>% 
  group_by(s_date,hour=hour(s_time)) %>% summarise(vals=n()) %>% arrange(s_date,desc(vals))

head(max_s_hour_data_for_max_date3)

sum(max_s_hour_data_for_max_date3$vals)

ggplot(max_s_hour_data_for_max_date3, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ s_date, ncol = 2) + 
  labs(
    title = "Count of Start Station 002587 Over Hours for Multiple Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()
```

```{r}
max_s_hour_data_for_max_date3_but3=jan_final_full%>% 
  filter(end_station_number=="002587",e_date %in% as.Date(c("2023-01-31","2023-01-12","2023-01-19","2023-01-26","2023-01-09","2023-01-11"))) %>%
  select(e_date,e_time,end_station_number) %>% 
  group_by(e_date,hour=hour(e_time)) %>% summarise(vals=n()) %>% arrange(e_date,desc(vals))

ggplot(max_s_hour_data_for_max_date3_but3, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ e_date, ncol = 2) + 
  labs(
    title = "Count of end Station 002587 Over Hours for above peak Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()
```


<span style="color: blue; font-size: 15px;"> 4)station 001075</span>

```{r 11}
head(station_data %>% filter(station_number=="001075") %>% arrange(desc(vals)))
max_s_hour_data_for_max_date4=jan_final_full%>% 
  filter(start_station_number=="001075",s_date %in% as.Date(c("2023-01-02","2023-01-01","2023-01-28","2023-01-30","2023-01-13","2023-01-21"))) %>%
  select(s_date,s_date,s_time,start_station_number) %>% 
  group_by(s_date,hour=hour(s_time)) %>% summarise(vals=n()) %>% arrange(s_date,desc(vals))

head(max_s_hour_data_for_max_date4)

sum(max_s_hour_data_for_max_date4$vals)

ggplot(max_s_hour_data_for_max_date4, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ s_date, ncol = 2) + 
  labs(
    title = "Count of Start Station 001075 Over Hours for Multiple Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()
```

```{r}
max_s_hour_data_for_max_date4_but4=jan_final_full%>% 
  filter(end_station_number=="001075",e_date %in% as.Date(c("2023-01-02","2023-01-01","2023-01-28","2023-01-30","2023-01-13","2023-01-21"))) %>%
  select(e_date,e_time,end_station_number) %>% 
  group_by(e_date,hour=hour(e_time)) %>% summarise(vals=n()) %>% arrange(e_date,desc(vals))

ggplot(max_s_hour_data_for_max_date4_but4, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ e_date, ncol = 2) + 
  labs(
    title = "Count of end Station 001075 Over Hours for above peak Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()
```


<span style="color: blue; font-size: 15px;"> 5)station 002696</span>

```{r 12}
head(station_data %>% filter(station_number=="002696") %>% arrange(desc(vals)))
max_s_hour_data_for_max_date5=jan_final_full%>% 
  filter(start_station_number=="002696",s_date %in% as.Date(c("2023-01-19","2023-01-09","2023-01-31","2023-01-30","2023-01-11","2023-01-26"))) %>%
  select(s_date,s_date,s_time,start_station_number) %>% 
  group_by(s_date,hour=hour(s_time)) %>% summarise(vals=n()) %>% arrange(s_date,desc(vals))

head(max_s_hour_data_for_max_date5)

sum(max_s_hour_data_for_max_date5$vals)

ggplot(max_s_hour_data_for_max_date5, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ s_date, ncol = 2) + 
  labs(
    title = "Count of Start Station 002696 Over Hours for Multiple Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()
```

```{r}
max_s_hour_data_for_max_date5_but5=jan_final_full%>% 
  filter(end_station_number=="002696",e_date %in% as.Date(c("2023-01-19","2023-01-09","2023-01-31","2023-01-30","2023-01-11","2023-01-26"))) %>%
  select(e_date,e_time,end_station_number) %>% 
  group_by(e_date,hour=hour(e_time)) %>% summarise(vals=n()) %>% arrange(e_date,desc(vals))

ggplot(max_s_hour_data_for_max_date5_but5, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ e_date, ncol = 2) + 
  labs(
    title = "Count of end Station 002696 Over Hours for above peak Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()

```


<span style="color: blue; font-size: 15px;"> overall all the top 5 start stations for their top 6 days hourly </span>



<span style="color: blue; font-size: 15px;">we can do a similar analysis for the end stations as well </span>


```{r 14, results='hide',warning=FALSE}
end_station_data=jan_final_full%>% 
  group_by(end_station_number) %>% 
  select(end_station_number) %>% 
  summarise(values=n()) %>% arrange(desc(values))

ggplot(end_station_data,aes(x=end_station_number,y=values))+geom_bar(stat="identity")
head(end_station_data)


top10_end_stations=head(end_station_data$end_station_number,10)
top10_end_stations

end_station_numbers <- c("001072","001011","000960","002587","001161")  


top_end_station_data_func <- function(data, station_number) {
  data %>%
    filter(end_station_number == station_number) %>%  
    group_by(e_date) %>%  
    summarise(vals = n(), .groups = 'drop') %>%  
    arrange(desc(vals)) %>%  
    mutate(
      weekday_or_weekend = ifelse(wday(e_date) %in% c(1, 7), "Weekend", "Weekday") 
    )
}


end_station_numbers <- c("001072","001011","000960","002587","001161")


end_station_data_list <- lapply(end_station_numbers, function(station) {
  top_end_station_data_func(jan_final_full, station)
})


end_station_data <- bind_rows(end_station_data_list, .id = "station_number")
end_station_data$station_number <- factor(end_station_data$station_number, labels = end_station_numbers)


head(end_station_data)
```

<span style="color: blue; font-size: 15px;"> top 5 end stations rental count for the january month </span>

```{r 15}
ggplot(end_station_data, aes(x = e_date, y = vals, fill = weekday_or_weekend)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ station_number, scales = "free") +
  labs(title = "Count of Start Stations Over Date", x = "Date", y = "Rented Bikes") +
  theme_minimal()

```

<span style="color: blue; font-size: 15px;"> top 10 end station over the heat map. </span>
```{r 15.1}
top_end_date_sattions=jan_final_full %>% filter(end_station_number %in% c("000959","000960","000961","000962","000963","000964","000965","000966", "000967","000968")) %>% 
  group_by(e_date,end_station_number) %>% summarise(count=n())

ggplot(top_end_date_sattions,
       aes(x = e_date, y = end_station_number, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "green") +
  labs(title = "Bike Rentals by Date and Start Station (Top 10 Stations)",
       x = "Date",
       y = "Start Station",
       fill = "count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```


<span style="color: blue; font-size: 15px;"> after identifying the top 5 station hourly rental graph for each station's top 6 days is drawn. </span>

```{r 16}

head(end_station_data %>% filter(station_number=="001072") %>% arrange(desc(vals)))
max_s_hour_data_for_max_end_date1=jan_final_full%>% 
  filter(end_station_number=="001072",e_date %in% as.Date(c("2023-01-24","2023-01-31","2023-01-09","2023-01-26","2023-01-12","2023-01-30"))) %>%
  select(e_date,e_time,end_station_number) %>% 
  group_by(e_date,hour=hour(e_time)) %>% summarise(vals=n()) %>% arrange(e_date,desc(vals))

#head(max_s_hour_data_for_max_end_date1)

ggplot(max_s_hour_data_for_max_end_date1, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ e_date, ncol = 2) +  # Facet by date, with 2 columns
  labs(
    title = "Count of End Station 001072 Over Hours for Multiple Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()

#head(end_station_data %>% filter(station_number=="001011") %>% arrange(desc(vals)))
max_s_hour_data_for_max_end_date2=jan_final_full%>% 
  filter(end_station_number=="001011",e_date %in% as.Date(c("2023-01-31","2023-01-12","2023-01-17","2023-01-24","2023-01-09","2023-01-11"))) %>%
  select(e_date,e_time,end_station_number) %>% 
  group_by(e_date,hour=hour(e_time)) %>% summarise(vals=n()) %>% arrange(e_date,desc(vals))

#head(max_s_hour_data_for_max_end_date2)

ggplot(max_s_hour_data_for_max_end_date2, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ e_date, ncol = 2) +  # Facet by date, with 2 columns
  labs(
    title = "Count of End Station 001011 Over Hours for Multiple Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()


#head(end_station_data %>% filter(station_number=="000960") %>% arrange(desc(vals)))
max_s_hour_data_for_max_end_date3=jan_final_full%>% 
  filter(end_station_number=="000960",e_date %in% as.Date(c("2023-01-13","2023-01-31","2023-01-11","2023-01-18","2023-01-27","2023-01-28"))) %>%
  select(e_date,e_time,end_station_number) %>% 
  group_by(e_date,hour=hour(e_time)) %>% summarise(vals=n()) %>% arrange(e_date,desc(vals))

#head(max_s_hour_data_for_max_end_date3)

ggplot(max_s_hour_data_for_max_end_date3, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ e_date, ncol = 2) +  # Facet by date, with 2 columns
  labs(
    title = "Count of End Station 000960 Over Hours for Multiple Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()


#head(end_station_data %>% filter(station_number=="002587") %>% arrange(desc(vals)))
max_s_hour_data_for_max_end_date4=jan_final_full%>% 
  filter(end_station_number=="002587",e_date %in% as.Date(c("2023-01-31","2023-01-12","2023-01-19","2023-01-26","2023-01-11","2023-01-09"))) %>%
  select(e_date,e_time,end_station_number) %>% 
  group_by(e_date,hour=hour(e_time)) %>% summarise(vals=n()) %>% arrange(e_date,desc(vals))

#head(max_s_hour_data_for_max_end_date4)

ggplot(max_s_hour_data_for_max_end_date4, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ e_date, ncol = 2) +  # Facet by date, with 2 columns
  labs(
    title = "Count of End Station 002587 Over Hours for Multiple Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()


#head(end_station_data %>% filter(station_number=="001161") %>% arrange(desc(vals)))
max_s_hour_data_for_max_end_date5=jan_final_full%>% 
  filter(end_station_number=="001161",e_date %in% as.Date(c("2023-01-13","2023-01-11","2023-01-12","2023-01-20","2023-01-30","2023-01-17"))) %>%
  select(e_date,e_time,end_station_number) %>% 
  group_by(e_date,hour=hour(e_time)) %>% summarise(vals=n()) %>% arrange(e_date,desc(vals))

#head(max_s_hour_data_for_max_end_date5)

ggplot(max_s_hour_data_for_max_end_date5, aes(x = hour, y = vals)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ e_date, ncol = 2) +  # Facet by date, with 2 columns
  labs(
    title = "Count of End Station 001161 Over Hours for Multiple Dates",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()

```


<span style="color: blue; font-size: 15px;"> from these graphs we can identify the startings and endings of bike trips. </span>


<span style="color: blue; font-size: 15px;"> we can do a analysis for same start end station </span>

```{r 17}
end_station_data=jan_final_full %>% select(end_station_number,end_station) %>% 
  group_by(end_station_number) %>% summarise(count=n())
#head(end_station_data)

equal_station_station=jan_final_full %>% filter(start_station_number==end_station_number,total_duration_mins>5) %>% 
  select(start_station_number,s_date,s_time,end_station_number,e_date,e_time,total_duration_mins) %>% 
  group_by(start_station_number) %>% summarise(count=n()) %>% arrange(desc(count))
head(equal_station_station)
p1=ggplot(equal_station_station,aes(x=start_station_number,y=count))+
  geom_bar(stat="identity",fill="lightblue")+
  labs(title = "same start and end station and their rental count",x="start station",y="no of rentals")
#p1


equal_station_hour=jan_final_full %>% filter(start_station_number==end_station_number,total_duration_mins>5) %>% 
  select(start_station_number,s_date,s_time,end_station_number,e_date,e_time,total_duration_mins) %>% 
  group_by(hour=hour(s_time)) %>% summarise(count=n())
#head(equal_station_hour)


equal_station_date=jan_final_full %>% filter(start_station_number==end_station_number,total_duration_mins>5) %>% 
  select(start_station_number,s_date,s_time,end_station_number,e_date,e_time,total_duration_mins) %>% 
  group_by(s_date) %>% summarise(count=n()) %>% arrange(desc(count))
#head(equal_station_date)
```



```{r 18}
top10__stations_dates_with_same_start_and_end=jan_final_full%>% 
  filter(start_station_number 
         %in% c("001075","300233","001157","001037","001132","200163","001088","001231","300227","001109"),
         s_date %in% as.Date(c("2023-01-02","2023-01-01","2023-01-28","2023-01-21","2023-01-15","2023-01-29","2023-01-22","2023-01-05","2023-01-06","2023-01-30")),
         start_station_number==end_station_number,
         total_duration_mins>5
  ) %>% 
  select(s_date,start_station_number) %>% 
  group_by(start_station_number,s_date) %>% 
  summarise(count=n()) %>% arrange(desc(count))
head(top10__stations_dates_with_same_start_and_end)

ggplot(top10__stations_dates_with_same_start_and_end,
       aes(x = s_date, y = start_station_number, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "green") +
  labs(title = "Bike Rentals by Date and Start Station for same start/end station (Top 10 Stations and top10 days)",
       x = "Date",
       y = "Start Station",
       fill = "count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

<span style="color: blue; font-size: 15px;"> check what happned in  station 300233 and 001075 on jan 02</span>

```{r 19}
same_start_end_station_300233_and_001075_jan_02=jan_final_full%>% 
  filter(start_station_number==end_station_number,total_duration_mins>5,s_date=="2023-01-02",
         start_station_number==c("300233","001075")) %>% 
  select(start_station_number,s_date,s_time) %>% 
  group_by(start_station_number,hour=hour(s_time)) %>% 
  summarise(count=n())


head(same_start_end_station_300233_and_001075_jan_02)

ggplot(same_start_end_station_300233_and_001075_jan_02, aes(x = hour, y = count)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ start_station_number, ncol = 2) +  
  labs(
    title = "Count of same Start Station and end station rentals Over Hours for the station 300233 and 001075 on jan 02",
    x = "Hour of the Day",
    y = "Rented Bikes"
  ) +
  theme_minimal()
```



<span style="color: black; font-size: 17px;">**3)Bike Model**.</span>

<span style="color: blue; font-size: 15px;">bike model is also an categorical variable,</span>

```{r 20}
bike_model_data=jan_final %>% group_by(bike_model) %>% select(bike_model) %>% 
  summarise(count=n())
head(bike_model_data)
ggplot(data=bike_model_data,aes(x=bike_model,y=count,fill=bike_model))+
  geom_bar(stat="identity",just=1)

```

<span style="color: red; font-size: 15px;">NOTE:we can clearly see that classic bikes are widely used more that E-bikes.</span>


<span style="color: black; font-size: 17px;">**4)BikeID**.</span>

<span style="color: blue; font-size: 15px;">this is also an categorical variable,</span>
 
```{r 21, warning=FALSE}
bikeID_data=jan_final %>% group_by(BikeID) %>% select(BikeID) %>% 
  summarise(count=n())
head(bikeID_data)
glimpse(bikeID_data)

ggplot(data=bikeID_data,aes(x=bikeID_data$BikeID,y=bikeID_data$count))+
  geom_bar(stat="identity",fill="lightblue")+
  labs(title = "bikeID and their rented counts",x="bikeID",y="count")

```

<span style="color: blue; font-size: 15px;">we can draw a boxplot identify outliers,</span>
```{r 22, warning=FALSE}
g3=ggplot(data=bikeID_data,aes(y=bikeID_data$count))+
  geom_boxplot(fill="lightyellow",color="yellow",outlier.color = "red",outlier.shape = 16)+
  labs(title = "boxplot of bikeID count",y="count")+
  theme_minimal()
g3
```

<span style="color: red; font-size: 15px;">NOTE:there are bikes that need to be removed because of too much usage.</span>

<span style="color: blue; font-size: 15px;">we can also draw a interactive boxplot,</span>

```{r 23, warning=FALSE}
ggplotly(g3)
```


<span style="color: black; font-size: 17px;">**5)Total duration in minutes**.</span>

<span style="color: blue; font-size: 15px;">this is the only numerical variable present in the corrected dataset,</span>

```{r 24}
ggplot(data = jan_final, aes(x = total_duration_mins)) +
  geom_density(fill = "lightblue", color = "black") +
  scale_x_log10() +  
  labs(title = "Density Plot of Total Duration (Minutes) - Log Scale", x = "Total Duration (Minutes) (Log Scale)", y = "Density")

ggplot(data = jan_final, aes(x = total_duration_mins)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 50) +
  scale_x_log10() + 
  labs(title = "Histogram of Total Duration (Mins) - Log Scale", x = "Total Duration (Mins) (Log Scale)", y = "Count")+
  geom_vline(aes(xintercept=mean(total_duration_mins)),color="red",linetype="dashed")+
  geom_vline(aes(xintercept=median(total_duration_mins)),color="blue",linetype="dashed")


mean(total_duration_mins)

```

<span style="color: red; font-size: 15px;">NOTE:as we can from the graph, we can assume that total duration is normally distributed with mean= 19.18 mins.</span>

```{r}
g4=ggplot(data = jan_final,aes(y=total_duration_mins))+
  geom_boxplot(color="green",fill="lightblue",outlier.color = "red")+scale_y_log10()

g4
```

<span style="color: blue; font-size: 15px;">from the boxplot we can identify the outliers and remove or fill them,</span>


<span style="color: blue; font-size: 15px;"> analyzing more about total duration</span>

```{r}
daily_total_duration=jan_final %>% select(s_date,total_duration_mins) %>% 
  group_by(s_date) %>% summarise(total_mins=sum(total_duration_mins))
head(daily_total_duration)

weekly_total_duration = daily_total_duration %>%
  mutate(week = cut(s_date, breaks = "week"))

# Boxplot by week 
ggplot(weekly_total_duration, aes(x = week, y = total_mins, fill = week)) +
  geom_boxplot() +
  labs(title = "Total Minutes by Week",x = "Week",y = "Total Minutes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```



```{r}

stationnwise_total_duration=jan_final %>% select(start_station_number,total_duration_mins) %>% 
  group_by(start_station_number) %>% summarise(total_mins=sum(total_duration_mins)) %>% 
  arrange(desc(total_mins))
#head(stationnwise_total_duration,10)

stationnwise_total_duration_top10_stations=jan_final %>% 
  select(start_station_number,total_duration_mins) %>% 
  filter(total_duration_mins>5,start_station_number 
         %in%c("001075","000960","200096","300221","001109","001011","001072","200128","300031","300083"))
#head(stationnwise_total_duration_top10_stations) 

ggplot(stationnwise_total_duration_top10_stations, aes(x = start_station_number, y = total_duration_mins, fill = start_station_number)) +
  geom_boxplot() +
  labs(title = "Total Minutes for top10 stations",x = "station",y = "Total Minutes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

stationnwise_total_duration_top10_stations %>%
  select(start_station_number,total_duration_mins) %>%
  group_by(start_station_number) %>% 
  summarise(total=sum(total_duration_mins),mean=mean(total_duration_mins),median=median(total_duration_mins),max=max(total_duration_mins),min=min(total_duration_mins))



stationnwise_total_duration_top5_dates=jan_final %>% 
  filter(start_station_number %in%c("001075","000960","200096","300221","001109")) %>% 
  select(start_station_number,s_date,total_duration_mins) %>% 
  group_by(start_station_number,s_date) %>% summarise(totalmins=sum(total_duration_mins))
head(stationnwise_total_duration_top5_dates)




ggplot(stationnwise_total_duration_top5_dates, aes(x = s_date, y = totalmins)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ start_station_number, ncol = 2) +  
  labs(
    title = "total duration of top 5 stations over the januray month",
    x = "total duration in mins",
    y = "date"
  ) +
  theme_minimal()


```


```{r}

stationnwise_total_duration_top5_hourly=jan_final %>% 
  filter(start_station_number %in%c("001075","000960","200096","300221","001109")) %>% 
  select(start_station_number,s_time,total_duration_mins) %>% 
  group_by(start_station_number,hour=hour(s_time)) %>% summarise(totalmins=sum(total_duration_mins)) %>% 
  arrange(desc(totalmins))
#head(stationnwise_total_duration_top5_hourly)

ggplot(stationnwise_total_duration_top5_hourly, aes(x = hour, y = totalmins)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  facet_wrap(~ start_station_number, ncol = 2) + 
  labs(
    title = "total duration of top 5 stations over the hours",
    x = "total duration in mins",
    y = "hour"
  ) +
  theme_minimal()
```

```{r}



top10_days=as.Date(c("2023-01-31" ,"2023-01-24" ,"2023-01-11", "2023-01-19" ,"2023-01-18" ,"2023-01-30", "2023-01-26", "2023-01-17","2023-01-12","2023-01-09"))
top10_stations=c("001072","001011","002587","001075","002696","300083","000960","001064","001142","022179")

top10_count_station_and_top10_days_and_totalduration=jan_final %>% 
  filter(start_station_number 
         %in% c("001072","001011","002587","001075","002696","300083","000960","001064","001142","022179"),
         s_date %in% as.Date(c("2023-01-31" ,"2023-01-24" ,"2023-01-11", "2023-01-19" ,"2023-01-18" ,"2023-01-30", "2023-01-26", "2023-01-17","2023-01-12","2023-01-09")),
         total_duration_mins>5
         ) %>% 
  select(s_date,start_station_number,total_duration_mins) %>% 
  group_by(start_station_number,s_date) %>% 
  summarise(totalmins=sum(total_duration_mins)) %>% 
  arrange(desc(totalmins))
#head(top10_count_station_and_top10_days_and_totalduration)

ggplot(top10_count_station_and_top10_days_and_totalduration,
       aes(x = s_date, y = start_station_number, fill = totalmins)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Bike Rentals by Date and Start Station (Top 10 Stations and top10 days)",
       x = "Date",
       y = "Start Station",
       fill = "total duration(in mins)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


```{r}
obviousmax1=top10_count_station_and_top10_days_and_totalduration %>% 
  filter(s_date=="2023-01-24",start_station_number=="001011")
head(obviousmax1)
obviousmax1_hourly=jan_final %>% 
  filter(s_date=="2023-01-24",start_station_number=="001011") %>% 
  select(start_station_number,s_date,s_time,total_duration_mins) %>% 
  group_by(hour=hour(s_time)) %>% summarise(totalmins=sum(total_duration_mins))
head(obviousmax1_hourly)
ggplot(obviousmax1_hourly,aes(x=hour,y=totalmins))+
  geom_bar(stat="identity",fill="lightblue")+
  labs(title = "hourly total duration in mins for 001011 for 2023-01-24",x="hour",y="total duration(in mins)")
obviousmax1_hourly %>% filter(hour==17)
```
```{r}
obviousmax1_count=jan_final %>% 
  filter(s_date=="2023-01-24",start_station_number=="001011") %>% 
  select(start_station_number,s_date,s_time,total_duration_mins) %>% 
group_by(hour=hour(s_time)) %>% summarise(count=n())

ggplot(obviousmax1_count,aes(x=hour,y=count))+
  geom_bar(stat="identity",fill="lightblue")+
  labs(title = "hourly count for 001011 for 2023-01-24",x="hour",y="count")
obviousmax1_count %>% filter(hour==17)
```



<span style="color: black; font-size: 17px;">**6)Start date**.</span>

<span style="color: blue; font-size: 15px;">This is also an categorical variable,</span>

```{r, warning=FALSE}
s_date_data=jan_final %>% group_by(s_date) %>% select(s_date) %>% summarise(count=n())
head(s_date_data)
ggplot(data=s_date_data,aes(x=s_date_data$s_date,y=s_date_data$count))+
  geom_bar(stat = "identity",fill="lightblue")+
  geom_smooth(method = "lm",se=FALSE)+
  labs(title = "start dates and count",x="date",y="count")

ggplot(data=s_date_data,aes(x=s_date,y=count))+geom_point(color="blue",size=3)
```

<span style="color: red; font-size: 15px;">Note:We can see that when reaching the end of the month there is trend of increasing no of rentals.</span>


```{r}
s_date_data_with_days=s_date_data %>% 
  mutate(
    weekday = wday(s_date, label = TRUE, abbr = FALSE), 
    weekday_or_weekend = ifelse(wday(s_date) %in% c(1, 7), "Weekend", "Weekday")  
  )
head(s_date_data_with_days)
attach(s_date_data_with_days)

ggplot(data=s_date_data_with_days,aes(x=s_date,y=count,color=weekday_or_weekend))+
  geom_point(size=3)
```

<span style="color: red; font-size: 15px;">NOTE:we can see that people are tend to rent bikes in weekdays more than weekends.</span>


<span style="color: black; font-size: 17px;">**7)Start Hour**.</span>

<span style="color: blue; font-size: 15px;">start hour is also an categorical variable,</span>

```{r, warning=FALSE}

hour_data=jan_final %>% group_by(hour=hour(jan_final$s_time)) %>% summarise(count=n())
head(hour_data)


ggplot(hour_data,aes(x=hour_data$hour,y=hour_data$count))+
  geom_bar(stat="identity",fill="lightblue")+
  labs(title = "hourly bike count",x="hour",y="bike count")+
  geom_smooth(method = "loess",color="red",se=FALSE,span=0.2)
```

<span style="color: red; font-size: 15px;">NOTE:we can see that there are peak hours.</span>

<span style="color: blue; font-size: 15px;">if we assume that hours which has bike count more that 20000 are peak hours,</span>

```{r, warning=FALSE}
ggplot(hour_data,aes(x=hour_data$hour,y=hour_data$count))+
  geom_bar(stat="identity",fill="lightblue")+
  labs(title = "hourly bike count",x="hour",y="bike count")+
  geom_smooth(method = "loess",color="red",se=FALSE,span=0.2)+
  geom_hline(aes(yintercept=23000),color="green",linetype="dashed",size=1.3)
```


### BIVARAINT ANALYSIS

<span style="color: black; font-size: 17px;">**1)start station and bike model**.</span>
```{r, warning=FALSE}
model_and_station=jan_final %>% group_by(bike_model,start_station) %>% summarise(count=n(),.groups = 'drop')
head(model_and_station)

ggplot(model_and_station,aes(x=model_and_station$start_station,
                             y=model_and_station$count,
                             color=model_and_station$bike_model)
       )+
  geom_point(size=0.6)
```

<span style="color: red; font-size: 15px;">NOTE:as we can see almost all the stations have a higher classic bike rentals.</span>


<span style="color: black; font-size: 17px;">**2)bike model total duration**.</span>
```{r}
model_and_duration=jan_final %>%select(bike_model,total_duration_mins) %>%  group_by(bike_model) %>%  summarise(count=n(),total_mins=sum(total_duration_mins))
head(model_and_duration)

ggplot(model_and_duration,aes(x=bike_model,y=total_mins,fill=bike_model))+geom_bar(stat="identity")

```

<span style="color: red; font-size: 15px;">NOTE:as we can see classic bike usage is far higher than E-bike usage.</span>


<span style="color: black; font-size: 17px;">**3)start station and total duration**.</span>
```{r}
station_and_duration=jan_final %>% select(start_station_number,total_duration_mins) %>% 
  group_by(start_station_number) %>% summarise(total_mins=sum(total_duration_mins))
head(station_and_duration)

ggplot(station_and_duration,aes(x=start_station_number,y=total_mins))+
  geom_bar(stat = "identity",fill="lightblue")
```


<span style="color: black; font-size: 17px;">**4)start date and total duration**.</span>
```{r,warning=FALSE}
date_and_duration=jan_final %>% select(s_date,total_duration_mins) %>% group_by(s_date) %>% 
  summarise(total_mins=sum(total_duration_mins))
head(date_and_duration)

ggplot(date_and_duration,aes(x=s_date,y=total_mins))+
  geom_bar(stat="identity",fill="lightblue")+
  geom_smooth(method = "loess",color="red",se=FALSE,span=0.1)+
  geom_smooth(method = "lm",color="green",se=FALSE,span=0.2)
```

<span style="color: blue; font-size: 15px;">including weekends and weekdays,</span>

```{r}
date_and_duration_weekends=date_and_duration %>% 
  mutate(weekday_or_weekend = ifelse(wday(s_date) %in% c(1, 7), "Weekend", "Weekday"))
head(date_and_duration_weekends)

ggplot(date_and_duration_weekends,aes(x=s_date,y=total_mins,fill=weekday_or_weekend))+
  geom_bar(stat="identity")
```

<span style="color: red; font-size: 15px;">NOTE:we can see that in saturdays bike usage is higher than sundays,but overall in weekdays bike usage is higher.</span>


<span style="color: black; font-size: 17px;">**5)start time and total duration**.</span>
```{r}
time_and_duration=jan_final %>% select(s_time,total_duration_mins) %>% 
  group_by(hour=hour(s_time)) %>% 
  summarise(total_mins=sum(total_duration_mins))
head(time_and_duration)
ggplot(time_and_duration,aes(x=hour,y=total_mins))+
  geom_bar(stat="identity",fill="lightblue")+
  geom_smooth(method = "loess",color="red",span=0.2,se=FALSE)
```

<span style="color: red; font-size: 15px;">NOTE:we can identify that from 6th hour to 19th bike usage is higher.</span>


<span style="color: black; font-size: 17px;">**6)bikeID and total duration**.</span>
```{r}
bikeID_and_duration=jan_final %>% select(BikeID,total_duration_mins) %>% 
  group_by(BikeID) %>% 
  summarise(total_mins=sum(total_duration_mins))
head(bikeID_and_duration)
ggplot(bikeID_and_duration,aes(x=BikeID,y=total_mins))+geom_bar(stat="identity",fill="red")

```

<span style="color: red; font-size: 15px;">NOTE:as we can see some bikes have higher usages.those bikes may need repairs or replacements.</span>


<span style="color: black; font-size: 17px;">**7)start time and start stations**.</span>
```{r}
time_and_stations=jan_final %>% select(s_time,start_station) %>% 
  group_by(hour=hour(s_time),start_station) %>% 
  summarise(count=n()) %>% arrange(desc(count))


unique_st=unique(time_and_stations$start_station)
set.seed(123)
random_st=sample(unique_st,10)
sample_time_and_station=time_and_stations %>% filter(start_station %in% random_st)
head(sample_time_and_station)

ggplot(sample_time_and_station, aes(x = hour, y = count, color = start_station)) +
  geom_line() +
  labs(title = "Bike Rentals Over hour by Start Station (random 10 Stations)",
       x = "hour",
       y = "Rental Count",
       color = "Start Station") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

<span style="color: red; font-size: 15px;">NOTE: we can see that for the random sample of 10 start stations,5-10 hours and 15-20 has a higher bike rentals.we can assume that they are the peak hours.</span>


<span style="color: black; font-size: 17px;">**8)start date and start station**.</span>
```{r}
date_stations=jan_final %>% select(s_date,start_station) %>% group_by(s_date,start_station) %>% 
  summarise(count=n())
head(date_stations)

unique_stations = unique(date_stations$start_station)
set.seed(123)  
random_stations = sample(unique_stations, 10)
filtered_data = date_stations %>%filter(start_station %in% random_stations)

ggplot(filtered_data, aes(x = s_date, y = count, color = start_station)) +
  geom_line() +
  labs(title = "Bike Rentals Over Time by Start Station (Top 10 Stations)",
       x = "Date",
       y = "Rental Count",
       color = "Start Station") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


ggplot(filtered_data, aes(x = s_date, y = start_station, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Bike Rentals by Date and Start Station (Top 10 Stations)",
       x = "Date",
       y = "Start Station",
       fill = "Rental Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

<span style="color: red; font-size: 15px;">NOTE: we can clearly see that when heading towards the ending of the January rental count gets higher.</span>


<span style="color: black; font-size: 17px;">**9)Bike model and start date**.</span>
```{r}
classic_bike_use=jan_final %>% filter(bike_model=="CLASSIC") %>% select(bike_model,s_date) %>%
  group_by(s_date) %>% summarise(c_count=n())
head(classic_bike_use)
attach(classic_bike_use)
ggplot(data=classic_bike_use,aes(x=s_date,y=c_count))+
  geom_bar(stat="identity",fill="lightblue",color="black")+
  geom_smooth(method = "loess", color = "red", se = FALSE, span = 0.2)+
  geom_smooth(method = "lm", color = "green", se = FALSE, span = 0.2)+
  labs(title = "classic bike count over time",x="start date",y="classic bike count")

Ebike_use=jan_final %>% filter(bike_model=="PBSC_EBIKE") %>% select(bike_model,s_date) %>%
  group_by(s_date) %>% summarise(e_count=n())
head(Ebike_use)
attach(Ebike_use)
ggplot(data=Ebike_use,aes(x=s_date,y=e_count))+
  geom_bar(stat="identity",fill="lightblue",color="black")+
  geom_smooth(method = "loess", color = "red", se = FALSE, span = 0.2)+
  geom_smooth(method = "lm", color = "green", se = FALSE, span = 0.2)+
  labs(title = "E-bike count over time",x="start date",y="E-bike count")

```

<span style="color: red; font-size: 15px;">NOTE:we can clearly see that both E-bikes and Classic bikes count increasing when heading towards the end of January.</span>


